---
title: "Sport Analytics Project"
output: word_document
date: "2024-12-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import Data
```{r}
# read in data
kickers <- read.csv("NFL_Kicker_Stats.csv")

# remove games when kickers were inactive, injured or other
kickers <- kickers[!kickers$Status %in% c("Inactive", "Injured Reserve", "COVID-19 List", "Exempt List"),]

# remove tied games
kickers <- kickers[kickers$Result != "T",]

# create binary variable for game win or loss
kickers$WinLoss <- ifelse(kickers$Result == "W", 1, 0)

# create continuous variable for kicker points missed 
kickers$PtsMissed <- (kickers$XPA - kickers$XPM) + 3*(kickers$FGA - kickers$FGM) 

# create continuous variable for difference in scores
kickers$ScoreDiff <- kickers$TmScore - kickers$OppScore

# create continuous variable for percentage of team's score that is kicker points
kickers$TmScoreKPts. <- ifelse(kickers$Pts == 0, 0, (kickers$Pts / kickers$TmScore) * 100)

# print first 6 rows
head(kickers)
```

Explain Data 
# All current NFL kickers (32) career statistics by game
# 23 variables with information about game and kicker statistics
# 2932 observations (every game a current kicker has played)

Inspiration
# The Steelers won against the Baltimore Ravens without scoring a touchdown, relying on six field goals from Boswell to get the 18-16 win
# It was Boswell's third career game with six FGs, which makes him the first kicker in NFL history to accomplish that

Problem Statement
# How important is a kicker to an NFL team? Do you need a successful kicker to win games? How detrimental is it to the team if a kicker misses field goals and extra points?
# Are we able to predict if a team will win a game based on game-by-game kicker statistics such as field goal percentages, extra point percentages, total number of points scored by the kicker?
# What relationships exist between kicker success and team success? Is there a noticeable difference in kicker performance for teams losing more games?

Highlight Important Variables
# Target: 
# WinLoss (0, 1) - whether team won or lost game

# Possible Predictors/Continuous Variables: 
# Player - kicker's name
# HomeAway - whether or not the game was home or away
# XP. - percentage of extra points kicker made out of total attempts
# FG. - percentage of field goals kicker made out of total attempts 
# Pts - total kicker points 
# PtsMissed - total points kicker failed to gain
# TmScoreKPts. - percentage of team's score that is kicker points
# ScoreDiff - difference in team's scores

Statistics to Look At
```{r}
# average difference in scores across these games
mean(abs(kickers$ScoreDiff))

# average percentage of kicker points that make up team's score
mean(kickers$TmScoreKPts.)

# average FG percentage
mean(kickers$FG.)

# average XP percentage
mean(kickers$XP.)

# probability of winning and losing from this data
sum(kickers$WinLoss == 1) / nrow(kickers)
sum(kickers$WinLoss == 0) / nrow(kickers)

# proportion of games won with a kicker that has a FG percentage of 80% or greater for that game
sum(kickers$FG. >= 80 & kickers$WinLoss == 1) / nrow(kickers)

# proportion of games won with a kicker that has a FG percentage less than 50% for that game
sum(kickers$FG. < 50 & kickers$WinLoss == 1) / nrow(kickers)

# proportion of games won with a kicker that has a XP percentage of 90% or greater for that game
sum(kickers$XP. >= 90 & kickers$WinLoss == 1) / nrow(kickers)

# proportion of games won with a kicker that has a XP percentage less than 60% for that game
sum(kickers$XP. < 60 & kickers$WinLoss == 1) / nrow(kickers)

# proportion of times a game was lost by less than or equal to the number of points missed by kicker
sum(abs(kickers$ScoreDiff) <= kickers$PtsMissed & kickers$Result == "L") / nrow(kickers)
```

Visuals (Charts, Graphs, Tables)
```{r}
# boxplot of kicker points and game results
boxplot(Pts ~ Result, data = kickers,
        main = "Kicker Points per Game vs Wins/Losses",
        xlab = "Wins vs Losses",
        ylab = "Kicker Points per Game",
        col = c("lightcoral","lightseagreen"))

# boxplot of kicker points missed and game results
boxplot(PtsMissed ~ Result, data = kickers,
        main = "Kicker Points Missed per Game vs Wins/Losses",
        xlab = "Wins vs Losses",
        ylab = "Kicker Points Missed per Game",
        col = c("lightcoral","lightseagreen"))

# histogram of percentage of score made up of kicker points
hist(kickers$TmScoreKPts.,
     main = "Histogram of Percentages of Team Score that is Kicker Points",
     xlab = "Percentage of Score that is Kicker Points",
     col = "lightblue")
# skewed to the right, but majority between 20-30%

# plot of wins by player
wins <- tapply(kickers$WinLoss, kickers$Player, sum)
sorted_wins <- sort(wins, decreasing = TRUE)
players <- names(sorted_wins)

plot(sorted_wins,
     main = "Number of Wins of Each Player in Career",
     xlab = "Player",
     ylab = "Number of Wins",
     xaxt = "n",
     col = "dodgerblue", 
     pch = 16,      
     las = 2)
axis(1, at = seq(1, 32, by = 1), labels = players1, las = 2, cex.axis = 0.5)

# plot of mean FG % by player
means <- tapply(kickers$FG., kickers$Player, mean)
sorted_means <- means[players]
plot(sorted_means,
     main = "Mean FG % of Each Player Over Career",
     xlab = "Player",
     ylab = "Mean FG %",
     xaxt = "n",
     col = "violet",
     pch = 16,      
     las = 2)  
axis(1, at = seq(1, 32, by = 1), labels = players, las = 2, cex.axis = 0.5)

# plot of mean XP % by player
means <- tapply(kickers$XP., kickers$Player, mean)
sorted_means <- means[players]
plot(sorted_means,
     main = "Mean XP % of Each Player Over Career",
     xlab = "Player",
     ylab = "Mean XP %",
     xaxt = "n",
     col = "violet",
     pch = 16,      
     las = 2)  
axis(1, at = seq(1, 32, by = 1), labels = players, las = 2, cex.axis = 0.5)
```

Observe Relationships between Variables (Tests)
```{r}
# t-test between kicker points in winning games v losing games
t.test(kickers$Pts[kickers$WinLoss == 1], kickers$Pts[kickers$WinLoss == 0])

# p-value is low, true difference in mean kicker points between wins and losses is not equal to 0
# kicker points are higher in games that are won

# t-test between kicker points missed in winning games v losing games
t.test(kickers$PtsMissed[kickers$WinLoss == 1], kickers$PtsMissed[kickers$WinLoss == 0])

# p-value is low, true difference in mean kicker points missed between wins and losses is not equal to 0
# less kicker points are missed in games that are won

#install.packages("vcd")
library(vcd)

# chisq-test between field goal % and win/loss
kicker_cat <- cut(kickers$FG., breaks = c(0, 50, 70, 90, 100), labels = c("Low", "Medium", "High", "Very High"))
winloss_fgp <- table(kickers$WinLoss, kicker_cat)
chisq.test(winloss_fgp)

# cramer's v
assoc_stats <- assocstats(winloss_fgp)
assoc_stats$cramer

# chisq-test between extra point % and win/loss
kicker_cat <- cut(kickers$XP., breaks = c(0, 50, 70, 90, 100), labels = c("Low", "Medium", "High", "Very High"))
winloss_xpp <- table(kickers$WinLoss, kicker_cat)
chisq.test(winloss_xpp)

# cramer's v
assoc_stats <- assocstats(winloss_xpp)
assoc_stats$cramer
```

Check for Multicollinearity
```{r}
# only consider Variables that are not highly correlated (cor > 0.6 or cor < -0.6)
cor_coeffs <- cor(kickers[,c(9,10,12:25)])
high_cor <- which(abs(cor_coeffs) > 0.6, arr.ind = TRUE)
high_cor <- high_cor[high_cor[,1] < high_cor[,2], ]
high_cor_values <- cor_coeffs[high_cor]
high_cor_vars <- data.frame(
  Var1 = colnames(cor_coeffs)[high_cor[, 1]],
  Var2 = colnames(cor_coeffs)[high_cor[, 2]],
  Correlation = high_cor_values
)
high_cor_vars
```

Models
```{r}
# predict team score for a game based on:
# extra point percentage, field goal percentage, kicker points, points missed by kicker
model1 <- lm(TmScore ~ XP. + FG. + Pts + PtsMissed, data = kickers)
summary(model1)
```

```{r}
# predict win or loss of game based on:
# kicker points, points missed by kicker
model2 <- glm(WinLoss ~ Pts + PtsMissed, data = kickers, family = "binomial")
summary(model2)
```
Pseudo R^2
```{r}
pseudo <- 1 - (model2$deviance / model2$null.deviance)
pseudo
```
Odds Ratios
```{r}
odds <- exp(coef(model2))
odds <- round(odds, 3)
odds

# for each additional kicker point, the odds of winning the game increase by about 26.2%
# for each additional kicker point missed, the odds of winning the game decrease by about 4.2%
```









